### 2020-02-10 paper/sex_selection_spacing-ver3.tex
Spacing discussion too diffuse and focused on detail

Understanding how birth spacing changes is essential for two main reasons.

First, if birth intervals increase substantially over time, India's fertility may 
be higher than generally accepted based on the total fertility rate (TFR).
Increasing birth spacing is a tempo effect, which, by postponing births, makes the TFR a 
downward biased estimate of cohort fertility 
\citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}.

Second, longer birth spacing is associated with lower mortality and 
morbidity risk \citep{Conde-Agudelo2012,Molitoris2019}.
In India, the duration to the next birth has traditionally been shorter with fewer sons, 
which likely contributed to the higher mortality and morbidity risk for girls
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.
Longer spacing between births then potentially lowers mortality, although there may
be counteracting effects if the longer spacing arises because of multiple abortions.
There has, so far, been no attempt to examine which mechanism dominates.

@@

I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
Using data from the four National Family and Health Surveys (NFHS), I apply
a competing risk hazard model to the birth histories of Hindu women, covering
the period 1972 to 2016.
The competing risk hazard model directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son, 
but also works in cases without prenatal sex selection.

% [Hypotheses / questions]
I address four main questions.

First, how did average birth intervals change over time for different groups, and how do 
these changes relate to the use of sex selection?

Birth intervals have increased for all education groups over the four decades, 
with larger increases, the higher the parity and the higher the education level.
Women who are most likely to use sex selection---well-educated women with no
sons---have seen the most substantial increases in birth intervals.
[TK check this!] As a result, well-educated women now show a reversal of the traditional spacing 
pattern, with the longest intervals for women with no sons.
Those least likely to use sex selection, women with no education in rural areas, 
still follow the traditional pattern of short spacing when they have girls and 
little evidence of the use of sex selection.

[Shorten/combine these two paragraphs]
Second, how has the distribution of birth spacing changed?
Average birth intervals help us understand the overall changes in 
spacing behavior, but tells us little about how these changes came about.
Examining how the distribution has change allow us to separate out increases in birth
spacing that come from women with strong son preference who still try to conceive early, 
but are also the most likely to end up with very long spells because of multiple 
abortions versus women who have longer average spacing because they try to avoid very
short birth intervals.

[TK this is unclear - complete rewrite]
The traditional pattern of longer average spacing with sons appears to arise from 
an increase in the number of very long intervals, rather than from fewer very short 
birth intervals. 
Across education groups, there appears to be little difference in the likelihood of 
having a very short birth interval across sex compositions, especially for lower parities.

Third, to what extent does TFR reflect cohort fertility?
The empirical model allows me to predict fertility behavior taking into
account both the likelihood of parity progression and having a child of a specific sex.
I compare those predictions with fertility rate estimates based on the same data to
show how the increases in birth spacing have led to a substantial overestimate of how 
fast fertility has declined. [TK too imprecise? underestimate of the level, rather?]

Finally, how has the relationship between birth spacing 
and mortality changed over time?
Mortality risk is inversely related to education level,
especially for very short birth intervals.
However, infant mortality risk has decline substantial 
over time for the lower education groups and is now
close to the level of women with the most education.
There is no evidence that the increasing use of 
sex selection is associated with higher mortality risk.

@@

The combination of rising incomes and continued son preference may even lead to longer 
spacing than what we would expect from the standard income effect alone.
As women's education increases, their productivity at home also increases, especially in 
the production of offspring human capital.
With relatively more boys born because of increased access to sex-selective 
abortions and an increasing potential income for (male) offspring, demand for 
better-educated women can increase, even if they do not participate in the labor market 
\citep{Behrman1999}.
The result would be longer birth intervals if parental attention per child results in 
higher child ``quality'' \citep{Zajonc1975,Zajonc1976,Razin1980}.
The evidence on spacing's effect on child quality measures such as IQ 
and education is, however, mixed for developed countries and non-existing 
for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.
Nevertheless, for one crucial aspect of child quality---health and mortality---longer 
spacing does lead to better outcomes, although this relationship weakens with maternal 
education \citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

@@

With the introduction of sex selection, parents can avoid the birth of girls, 
although at the cost of the procedure and having to wait longer until the next child 
for each aborted pregnancy.
Theory suggests that sex selection increases with lower desired fertility, and, for a 
given desired number of children, the higher the parity \citep{Portner2015b}.
The increased use of sex selection in India with education and in urban
areas is consistent with a lower desired fertility for both groups
\citep{das_gupta97,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Better-educated women also tend to live in households with higher income, which better 
enables them to access prenatal sex determination and lowers the relative costs of using 
sex selection and have long birth intervals.
We may even observe a ``delay'' effect, where parents conceive later because they know 
that short spacing is detrimental to the next child's health, and that the next child is 
more likely to be a boy because of access to sex selection.

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect a push toward longer birth spacing over time, independent
of education levels, based on the income effects and the effects of spacing
on child outcomes.
As desired fertility decreases with increasing education, I, furthermore, expect 
birth spacing to increase the most among the better-educated because of their 
use of sex selection.%
\footnote{
Working in the opposite direction is that women with more education can space
their children closer together without substantially increasing child mortality risk,
although compared with those factors this effect is likely to be relatively small.
}
Even with the large increase in the number of better-educated women, "Sanskritization" 
imply that the changing composition does not substantially change this group's use of 
sex selection.

@@

Second, it is unlikely---even in the absence of prenatal sex determination---that 
characteristics, such as the sex composition of previous births, have the same effects 
throughout the entire spell as is assumed by the proportional hazard model.
The proportionality assumption is especially problematic for higher-order spells where 
there are substantial differences across groups both in the likelihood of progressing to 
the next birth and how soon couples want their next child if they are going to have one.
The introduction of prenatal sex determination exacerbates any bias from the 
proportionality assumption since sex-selective abortion use varies across groups,
which affects birth spacing, and because a household's use of sex selection may vary 
within a spell, which means that the effects of covariates vary within the spell as well.
I, therefore, use a non-proportional hazard specification, which allows the shape of the 
hazard functions to vary across groups.
The use of a non-proportional specification also mitigates any potential effects 
of unobserved heterogeneity when used in conjunction with a flexible baseline hazard 
\citep{Dolton1995}.

@@

% [How to read tables]
I estimate the model for each spell/education/period subsample, the results of which I use 
to predict average birth spacing, sex ratio, and the probability of having a birth in that 
spell.
Figures \ref{fig:bootstrap_low} through \ref{fig:bootstrap_highest}
show these predicted outcomes for the four education levels separated by the area of 
residence.%
\footnote{
For legibility, the graphs do not show standard errors.
The graphed values, with standard errors, are shown in Appendix Tables 
\ref{tab:avg_sex_ratio_low} through \ref{tab:avg_sex_ratio_highest}.
}

@@

\subsection{No Education Women}

@@

\captionsetup[subfigure]{captionskip=-1pt,farskip=-0.5pt}

@@

\begin{sidewaysfigure}
\centering
\rotatebox[origin=c]{90}{\small{Urban}}
\subfloat[Second Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell2_low_urban_all}
    \end{minipage}
} 
\subfloat[Third Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell3_low_urban_all} 
    \end{minipage}
}
\subfloat[Fourth Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell4_low_urban_all}
    \end{minipage}
}
\caption{Estimated average spell length, sex ratio, and probability of 
a next birth for women with no education}
\label{fig:bootstrap_low}
\end{sidewaysfigure}

@@

\subsection{Low to Medium Education Women}

@@

@@

\subsection{High Education Women}

@@

\subsection{Highest Education Women}

The impact that sex selection can have on birth spacing is illustrated particularly 
vividly by the reversal of the standard spacing pattern for the third spell for urban 
women with twelve or more years of education as shown in 
Figure \ref{fig:bootstrap_highest}.%
\footnote{
I do not show the results for the fourth spell since the low number of births,
especially for women with one or more boys make the spacing and sex ratio results
very noisy.
Results are available upon request.
}
In the 1972-1984 period, the predicted average birth intervals for the three possible sex 
compositions were only a couple of months apart, with the shortest spacing for women who 
had two girls.
Women with two girls had, however, the highest predicted average spacing in the later
periods, with about seven months longer birth intervals than women with one or two boys.

@@

Consistent with a larger bias in the fertility rate when age of marriage and the length of 
birth intervals are increasing, the absolute bias is least in the first and the last 
period and highest in the middle two periods.
Hence, the fertility rate declined too fast between the mid-1980s and the end of the 
century, and only recently as the rate of increase for the birth intervals have slowed 
are the predicted cohort fertility and the fertility beginning to converge again.
Even with the convergence, the predicted cohort fertility for the last period is above 
the 1992--93 fertility rate for every group except urban women with no education.
Furthermore, for the last period, the predicted cohort fertility is stil between
and 20 percent higher than the fertility rate.

>>>>

Understanding how birth spacing changes is of interest for two main reasons, in 
addition to a better general understanding how fertility decisions are made.

First, if birth intervals increased substantially over time, India's fertility may 
be higher than generally accepted based on the total fertility rate (TFR).
Increasing birth spacing is a tempo effect, which, by postponing births, makes the TFR a 
downward biased estimate of cohort fertility 
\citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}.

Second, longer birth spacing is associated with lower mortality and 
morbidity risk \citep{Conde-Agudelo2012,Molitoris2019}.
In India, the duration to the next birth has traditionally been shorter with fewer sons, 
which likely contributed to the higher mortality and morbidity risk for girls
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.
Longer spacing between births then potentially lowers mortality, although there may
be counteracting effects if the longer spacing arises because of multiple abortions.
There has, so far, been no attempt to examine how mortality has responded to the
increase in sex selection.

@@

I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
Using data from the four National Family and Health Surveys (NFHS), I apply
a competing risk hazard model to the birth histories of Hindu women, covering
the period 1972 to 2016.
The competing risk hazard model directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son, 
but also works in cases without prenatal sex selection.
The empirical model allows me to predict fertility behavior taking into
account both the likelihood of parity progression and having a child of a specific sex.

% [Hypotheses / questions]
There are four main set of results.

First, birth intervals have increased for all education groups over the four decades, 
with larger increases, the higher the parity and the higher the education level.
Women who are most likely to use sex selection---well-educated women with no
sons---have seen the most substantial increases in birth intervals.
[TK check this] As a result, well-educated women now show a reversal of the traditional spacing 
pattern, with the longest intervals for women with no sons.
Those least likely to use sex selection, women with less education in rural areas, 
still follow the traditional pattern of short spacing when they have girls and 
little evidence of the use of sex selection.
[TK add sex ratios and parity progression discussions]

Second, there is little difference over time and across groups in the likelihood of 
having a very short birth interval given the sex composition of prior children, 
especially for the lower parities.
Hence, the increases in average spacing come predominantly from the longer birth intervals
getting even longer over time, driven to a large degree by a substantial use of sex 
selection.
The exception to this result is that the better-educated women with \emph{very high} 
predicted use of sex selection appear to conceive later exactly because the next child is 
almost certainly a boy.

Third, the fertility rate substantially overestimated how fast cohort fertility fell 
in the 1990s and early 2000s, when spacing began to increase as the result of secular 
changes and increasing use of sex selection.
Although predicted cohort fertility and the fertility rate has lately begun to converge,
the predicted cohort fertility is still ten to twenty percent higher than the
fertility rate.
Furthermore, even with the convergence, predicted cohort fertility is still at or above 
replacement level, except for the best-educated urban women.

Finally, infant mortality risk has decline substantially over time for all groups,
but fastest for the lower education groups, who are now close to the level of women with 
the most education.
Mortality risk is, however, still inversely related to education level, especially for 
very short birth intervals.
There is no evidence that the increasing use of sex selection is associated with higher 
mortality risk.

@@

The combination of rising incomes and continued son preference may lead to even longer 
spacing than the standard income effect alone predicts.
As women's education increases, their productivity in the production of offspring human 
capital also increases.
With relatively more boys born because of increased access to sex-selective 
abortions and an increasing potential income for (male) offspring, demand for 
better-educated women can increase, even if they do not participate in the labor market 
\citep{Behrman1999}.
If more and ``better'' parental attention per child results in higher child ``quality'' 
we should expect longer birth intervals 
\citep{Zajonc1975,Zajonc1976,Razin1980}.
The evidence on spacing's effect on child quality measures such as IQ 
and education is, however, mixed for developed countries and non-existing 
for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.
The exception is one crucial aspect of child quality---health and mortality---where
longer spacing does lead to better outcomes, although this relationship weakens with 
maternal education \citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

@@

The introduction of sex selection allows parents to avoid the birth of girls,
but increases the expected duration to the next birth.
Theory suggests that sex selection increases with lower desired fertility, and, for a 
given desired number of children, the higher the parity \citep{Portner2015b}.
The increased use of sex selection in India with education and in urban
areas is consistent with a lower desired fertility for both groups
\citep{das_gupta97,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Better-educated women also tend to live in households with higher income, which better 
enables them to access prenatal sex determination and lowers the relative costs of using 
sex selection and have long birth intervals.

We may even observe a ``delay'' effect, where parents conceive later because they know 
that short spacing is detrimental to the next child's health, and that the next child is 
more likely to be a boy because of access to sex selection.
Working in the opposite direction is that women with more education can, in principle, 
space their children closer together without substantially increasing child mortality 
risk.
Given the apparent lack of pressure to return to the labor force,
this effect is, however, likely to be small.

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect a push toward longer birth spacing over time, independent
of education levels, based on the income effects and the effects of spacing
on child outcomes.
As desired fertility decreases with increasing education, I, furthermore, expect 
birth spacing to increase the most among the better-educated because of the larger
increase in household income income---even with declining female labor participation---and 
their use of sex selection.
Even with the large increase in the number of better-educated women, "Sanskritization" 
imply that the changing composition does not substantially change this group's use of 
sex selection.

@@

Second, it is unlikely---even in the absence of prenatal sex determination---that 
characteristics, such as the sex composition of previous births, have the same effects 
throughout the entire spell as is assumed by the proportional hazard model.
The proportionality assumption is especially problematic for higher-order spells where 
there are substantial differences across groups both in the likelihood of progressing to 
the next birth and how soon couples want their next child if they are going to have one.

The introduction of prenatal sex determination exacerbates any bias from the 
proportionality assumption since sex-selective abortion use varies across groups,
which affects birth spacing, and because a household's use of sex selection may vary 
within a spell, which means that the effects of covariates vary within the spell as well.

I, therefore, use a non-proportional hazard specification, which allows the shape of the 
hazard functions to vary across groups.
The use of a non-proportional specification also mitigates any potential effects 
of unobserved heterogeneity when used in conjunction with a flexible baseline hazard 
\citep{Dolton1995}.

@@

% [How to read tables]
I estimate the model for each spell/education/period subsample, the results of which I use 
to predict average birth spacing, sex ratio, and the probability of having a birth in that 
spell.
Figures \ref{fig:bootstrap_low} through \ref{fig:bootstrap_highest}
show these predicted outcomes for the four education levels separated by the area of 
residence.%
\footnote{
For legibility, the graphs do not show standard errors.
The graphed values, with standard errors, are shown in Appendix Tables 
\ref{tab:avg_sex_ratio_low} through \ref{tab:avg_sex_ratio_highest}.
Furthermore, I do not show the results for the fourth spell the best-educated women since 
the low number of births, especially with one or more boys, make the spacing and sex ratio 
results very noisy.
Results are available upon request.
}

@@

@@

@@

\begin{sidewaysfigure}
\captionsetup[subfigure]{captionskip=-1pt,farskip=-0.5pt}
\centering
\rotatebox[origin=c]{90}{\small{Urban}}
\subfloat[Second Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell2_low_urban_all}
    \end{minipage}
} 
\subfloat[Third Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell3_low_urban_all} 
    \end{minipage}
}
\subfloat[Fourth Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell4_low_urban_all}
    \end{minipage}
}
\caption{Estimated average spell length, sex ratio, and probability of 
a next birth for women with no education}
\label{fig:bootstrap_low}
\end{sidewaysfigure}

@@

@@

Average birth intervals have increased over time across all groups, and are, for a given 
parity, longer the more education a woman.
The increases in birth spacing are larger the higher the parity
and, within parities, the higher the education.

In the latest period, the lowest average birth interval is for women with no 
education in the second spell at just shy of 25 months, which is an increase of a
couple of months from the first period.
The longest average interval was 42 months for the third spell for women with twelve or 
more years of education who had no boys, which is an increase of approximately a year.
For comparison, the median birth intervals reported for the NFHS have barely
moved over time, staying at approximately 22 months calculated with the starting point 
nine months after the prior birth used here.%
\footnote{
See
\citet[][p.\ 110--112]{International-Institute-for-Population-Sciences-IIPS1995},
\citet[][p.\ 98--103]{International-Institute-for-Population-Sciences-IIPS2000},
\citet[][p.\ 88--91]{International-Institute-for-Population-Sciences-IIPS2007}, and
\citet[][p.\ 81--82]{International-Institute-for-Population-Sciences-IIPS2017}.
Median estimates are in
Appendix Tables \ref{tab:p25_p50_p75_low} through \ref{tab:p25_p50_p75_highest}.
}
This lack of change underscores the importance of accounting for censoring of birth spells 
when trying to understand how birth spacing is changing over time.

Parity progression probabilities are lower the higher the education and are
falling across all groups over time, but the extent of the decline depends crucially on 
the sex composition of prior children and the parity.
As expected with a strong son preference, women with no sons still are the most likely to 
have a next birth, and this group has seen often substantially smaller declines in
the likelihood of having another birth than women with one or more sons.%
\footnote{
There is, however, also some evidence that parents prefer a mixed sex composition: for 
the fourth spell women with two boys \emph{and} a girl
consistently have the lowest parity progression probability.
}
For urban women with eight to eleven years of education, for example, the second spell
probability of a next child declined by eight percentage points if the first child was a 
girl, but 15 percentage points if it was a boy.
Even more extreme is the fourth spell where the likelihood of a birth declined by six 
percentage points to 75 percent if the woman had three children, but by almost 40 
percentage points to 25 percent if she had two boys and a girl.

To what extent has the introduction of sex selection changed birth spacing across sex 
compositions?
There are three broad groups in the data.

First, women who have only girls but do not use sex selection had, and continue to have,
the shortest spacing, as shown, for example, by rural women with no education.
For this group spacing is shortest with only girls across all spell, and the difference
between sex compositions have even widen over time.

Second, even with sex selection, and the associated increases in expected spacing, the
intervals for the other sex compositions have increased even more and the relatively 
pattern, therefore, remains more or less the same.
This is the case, for example, for the second spell for women with eight to eleven
and twelve or more years of education.

Finally, women with the most unequal sex ratios show the longest spacing in the 
absence of boys, and this spacing is longer than for the other possible sex 
compositions.
The most extreme example is that with a sex ratio for the best-educated urban women  
approaching 80 percent, the 42 months birth intervals with two girls is seven months 
longer than with at least one boy.
Even urban women with eight to eleven years of education show the longest intervals with 
only girls in the last period.

@@

@@

The impact that sex selection can have on birth spacing is illustrated particularly 
vividly by the reversal of the standard spacing pattern for the third spell for urban 
women with twelve or more years of education as shown in 
Figure \ref{fig:bootstrap_highest}.
In the 1972-1984 period, the predicted average birth intervals for the three possible sex 
compositions were only a couple of months apart, with the shortest spacing for women who 
had two girls.
Women with two girls had, however, the highest predicted average spacing in the later
periods, with about seven months longer birth intervals than women with one or two boys.

@@

Consistent with a larger bias in the fertility rate when age of marriage and the length of 
birth intervals are increasing, the absolute bias is least in the first and the last 
period and highest in the middle two periods.
Hence, the fertility rate declined too fast between the mid-1980s and the end of the 
century, and only recently as the rate of increase for the birth intervals have slowed 
are the predicted cohort fertility and the fertility beginning to converge again.
Even with the convergence, the predicted cohort fertility for the last period is above 
the 1992--93 fertility rate for every group except urban women with no education.
Furthermore, for the last period, the predicted cohort fertility is stil between ten
and 20 percent higher than the fertility rate.
